Poker Chips, Earthquakes, and Computation

### 2014-09-27 draft0.tex
Edited Section "Acyclitiy"
Changed the denominator from 4(|x|+|y|) to 4(|x|+|y|+1) so that the function does not blow up for (0,0). Also added structure to make more similar to previous sections

%%This section needs to be fixed up

It turns out that a potential function argument can also show that infinite boards with a bounded number of chips on each square can never cycle.

Consider the following potential function
\[ \Phi(f) = \sum_{x,y}\frac{f(x,y)}{4(|x|+|y|)\cdot4^{(|x|+|y|)}} \]
First, we note that for $n$, there are exactly $4n$ squares $(x,y)$ such that $|x| + |y| = n$. If we assume that the maximum chip stack is $M$, then the potential function can be bounded:
\begin{align*}
\sum_{x,y}\frac{f(x,y)}{4(|x|+|y|)\cdot4^{(|x|+|y|)}} \leq \sum_{n=1}^\infty \frac{M}{4^n} = \frac{M}{3}
\end{align*}
If any square at distance $d$ is hit by a local earthquake, the value of the potential at that square decreases by $\frac{4}{4d\cdot 4^d}$. The potential increases on the four adjacent squares. In particular we know that one of the adjacent squares is at distance $d-1$, and has a potential increase of $\frac{1}{4(d-1)\cdot 4^{d-1}} = \frac{4}{4(d-1)\cdot 4^d}$ which is larger in magnitude than $\frac{4}{4d\cdot 4^d}$. So the potential strictly increases after any local earthquake.

It follows that infinite boards hit with local earthquakes can never cycle back to a previous arrangement

>>>>

We can in fact show that an infinite board with a bounded number of chips on each square cannot cycle under local earthquakes. To show this we will mirror the previous section by defining a potential function on boards and then showing that the functions value only increases after a local earthquake has occurred.

We make the quick definition of the set of infinite boards
\begin{definition}
We say that a board is infinite if the total number of chips on it is finite. That is a board $f$ is infinite if the sum 
\begin{equation*}
 \sum_{(x,y) \in \mathbb{Z} \times \mathbb{Z}} f(x,y)
\end{equation*}
does not converge. We denote the set of infinite boards as $\mathcal{B}_I$. We say that an infinite board $f \in \mathcal{B}_I$ is bounded if $\exists M \in \mathbb{Z}$ such that $f(x,y) \leq M$ for all $(x,y) \in \mathbb{Z} \times \mathbb{Z}$.  We denote the set of bounded infinite boards as $\mathcal{B}_B$
\end{definition}

We once again define a potential function, although this time it is from bounded infinite boards to the positive reals. \begin{definition} Let $\Phi: \mathcal{B}_B \rightarrow \mathbb{R}_{\geq 0}$
\[ \Phi(f) = \sum_{(x,y)}\frac{f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} \]
\end{definition}
%include the +1 in the denominator so there is not a singularity at (0,0)
Before it was obvious that our potential function converged and was well defined. However we are now considering boards for which the sum of all chips does not converge. Therefore we show 
\begin{lemma} $\Phi(f)$ converges for all $f \in \mathcal{B}_B$.
\end{lemma}
\begin{proof}
We note that for $n > 0$, there are exactly $4n$ squares $(x,y)$ such that $|x| + |y| = n$. If we assume that the maximum chip stack is $M$, then we have the bound:
\begin{align*}
\Phi(f) = \sum_{(x,y)}\frac{f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} \leq \sum_{n=1}^\infty \frac{M}{4^n} = \frac{M}{3}
\end{align*}
Thus the sum in the potential function converges and the function is well defined. 
\end{proof}

\begin{theorem} 
If $f \in \mathcal{B}_B$ and $LE(f, (x_0,y_0)) \neq f$, then $LE^{n}(f) \neq f$ for all $n>0$.
\end{theorem}
\begin{proof}
We once again compute 
\begin{align*}
\Phi(LE(f))-\Phi(f) = \sum_{(x,y)} \frac{LE(f, (x_0,y_0))(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} - \frac{f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} \\
=\frac{LE(f, (x_0,y_0))(x_0,y_0)-f(x_0,y_0)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} + \sum_{(x,y) \in n(x_0,y_0)} \frac{LE(f, (x,y))(x_0,y_0)-f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} 
\end{align*}
Suppose $|x|+|y|=d$, then 
\begin{equation*}
\frac{LE(f,(x,y))(x,y)}{4(d+1)*4^d} -\frac{f(x,y)}{4(d+1)*4^d} =-\frac{4}{4(d+1)\cdot 4^d}
\end{equation*}
and since $LE(f, (x,y))(x_0,y_0)>f(x,y)$ for $(x,y) \in n(x_0,y_0)$ we have
\begin{equation*}
\frac{LE(f, (x,y))(x_0,y_0)-f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}} >0
\end{equation*}
Furthermore we know that one of the adjacent squares is at distance $d-1$, and that 
\begin{equation*}
\frac{LE(f, (x,y))(x_0,y_0)-f(x,y)}{4(|x|+|y|+1)\cdot4^{(|x|+|y|)}}  = \frac{1}{4(d)\cdot 4^{d-1}} = \frac{4}{4(d)\cdot 4^d}
\end{equation*}which is larger in magnitude than $\frac{4}{4(d+1)\cdot 4^d}$. Therefore the value of $\Phi$ strictly increases after any local earthquake.
It follows that infinite boards hit with local earthquakes can never cycle back to a previous arrangement.
\end{proof}
